---
title: "Course Project"
author: "Subha"
date: "July 21, 2015"
output: html_document
---

These are the files produced during a homework assignment of Coursera’s MOOC Practical Machine Learning from Johns Hopkins University. 
The scripts tested and executed on MAC OS X 10.10.2, and RStudio Version 0.98.1103


### Loading the libraries needed for the project
```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
```
### Downloading the training and test files for this assignment
```{r,echo=TRUE}
fileURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(fileURL,destfile="./training_data.csv",method="curl")

## reading the training file to R
training <- read.csv("./training_data.csv",header=T)

## downloading and reading the test file into R
fileURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(fileURL,destfile="./test_data.csv",method="curl")
testing <- read.csv("./test_data.csv",header=T)

```
### Cleaning Data for analysis 
```{r,echo=TRUE}
##No of rows and columns in training
dim(training)
## removing columns that have "NA" from training
new_data <- training[ , ! apply( training , 2 , function(x) any(is.na(x)) ) ]
## removing the timestamp columns and sequence data from new data set
new_data<-new_data[,-c(1:7)]
## Checking for near zero variables
non_zero <- nearZeroVar(new_data, saveMetrics=FALSE)
print(non_zero)

## using only columns that do not have non zero variance for our final predictions
new_data <- new_data[,-(non_zero)]
dim(new_data)
```{r}
```
### Reproduceability 
```{r}
set.seed(100)

## Dividing the data into training and test
inTrain <- createDataPartition(new_data$classe,p=0.7,list=F)
train_set <- new_data[inTrain,]
test_set <- new_data[-inTrain,]

## Ensuring the testing data set provided also has the same no of columns as the training set
## removing the classe column
```{r,echo=TRUE}
new_col_list <- colnames(new_data[,-53])
testing <- testing[new_col_list]
```{r}
```
### ML algortihms for prediction
```{r,echo=TRUE}
##Model Building : Using "rpart" - decision tree algorithm
modFit<-train(classe~.,method = "rpart",data=train_set)
fancyRpartPlot(modFit$finalModel)
## Predicting...
pred_rpart<-predict(modFit,test_set)
## Confusion Matrix to test and report the results
results<-confusionMatrix(pred_rpart,test_set$classe)
print(results)
```{r}
```
##Analysis
### As accuracy is very low with rpart, lets try building with randomForest
```{r,echo=T}
## Model Building using Random forest algorithm
model_rf <- randomForest(classe~.,data=train_set)
##Predictions:
pred_rf <-predict(model_rf,test_set)
## ConfusionMatrix to publish results
results_rf<-confusionMatrix(pred_rf,test_set$classe)
print(results_rf)
##Accuracy is > 99% with Random Forest. We will use this model to predict against the test
## set of 20 observations provided for testing our algorithm
```
###Final Predictions with our model
```{r,echo=TRUE}
predictTest <- predict(model_rf,testing)
## print the results of prediction
print(predictTest)
```{r}
```
## Out of Sample Error
```{r}
### out of sample error is the error rate we get on a new data set.
### I am using the RandomForest algorithm for my predictions and can say the ### average out of sample error will be 

##Random Forest Testing Set: 1 - .9954 = 0.0046
```{r}
```
### Validating my estimate out of error by cross validation
```{r,echo=TRUE}
modFit_cv <-randomForest(classe~., data=train_set,method="cv")
pred_cv <- predict(modFit_cv,test_set)
confusionMatrix(pred_cv,test_set$classe)
```{r}
### Out of sample error = 1 - Accuracy ( 1 - .9959 = 0051)
```{r}

```
Function to predict the answers
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
```
Output the answers
```{r,echo=TRUE}
answers = predictTest
pml_write_files(answers)


